---
title: "flashfm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flashfm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Flashfm (flexible and shared information fine-mapping) is a package to simultaneously fine-map genetic associations for
multiple quantitative traits that are measured in the same study(s) by sharing information between the traits. It is flexible 
to the inclusion of related individuals in the sample and to missing trait measurements.

Flashfm makes use of summary-level data and requires as input: 
 
- GWAS summary statistics  
- trait covariance matrix 
- vector of trait means 
- reference genotype matrix OR reference genotype covariance matrix and vector of RAFs 
- single-trait fine-mapping results (models and their posterior probabilities); within flashfm is a function for single-trait fine-mapping, 
if needed 

This vignette introduces flashfm and gives an illustration of its use - with both forms of input for the reference genotype information - 
on simulated data of two traits with a shared causal variant. A simulated data set is provided in flashfm.

# Flashfm overview

Flashfm shares information between traits by up-weighting joint models that have a shared causal variant between traits, in a similar 
Bayesian framework to Multinomial Fine-Mapping ([MFM](https://www.nature.com/articles/s41467-019-11271-0)) for joint fine-mapping of multiple 
diseases with shared controls.

To generate posterior support for fine-mapping models, flashfm needs to calculate the Bayes' factor (BF) for all possible model 
combinations across SNPs and traits. To make this computationally feasible, we derive an expression for the the joint BF that depends on 
the individual (marginal) trait BFs and terms that depend on the GWAS summary statistics, covariance matrix of the traits, and sample 
sizes; either a genotype matrix or both a SNP covariance matrix and relative allele frequencies (RAFs) are needed from a reference panel or 
in-study sample.

As in MFM, rather than reporting posterior probabilities (PPs) for each SNP model, we focus on SNP groups that we construct such that 
SNPs in the same group are in LD, have a similar effect on traits, and rarely appear together in a trait model. We constuct SNP groups 
based on the model PPs from single-trait fine-mapping and a separate set of SNP groups based on the model PPs from flashfm. In general, 
the groups constructed by flashfm tend to be a subset of those from single-trait fine-mapping, indicating that flashfm gives finer 
resolution that single-trait fine-mapping.

Details of how to run flashfm on your own data are explained through detailed examples of simulated data. 
We provide three ways to run flashfm, depending on whether or not single-trait fine-mapping results are already available or not.

1. Single-trait fine-mapping has not yet been run - need GWAS summary statistics for each trait, SNP correlation matrix, 
reference allele frequency vector, and trait covariance matrix. In one command, run single-trait fine-mapping and flashfm, as well as construct SNP groups 
for each approach and summarise results by SNPs and by SNP groups. See first example.

2. Single-trait fine-mapping has not yet been run - need GWAS summary statistics for each trait, SNP matrix, and 
trait covariance matrix. See second example.

3. Single-trait fine-mapping has been run using [FINEMAP](https://academic.oup.com/bioinformatics/article/32/10/1493/1743040) 
- need *config files from FINEMAP for each trait, GWAS summary statistics for each trait, SNP correlation matrix, 
reference allele frequency vector, and trait covariance matrix. See third (last) example.

# Simulation Example

In this simulated data example, we simulate two traits that each have two causal variants, of which one (rs61839660) is shared between 
traits; trait 1 has second causal variant rs62626317 and trait 2 has second causal variant rs11594656. The trait correlatin is 0.4.

For a region of 345 SNPs in chromosome 10p-6030000-6220000 (GRCh37/hg19), containing IL2RA, 
we generated a population of 100,000 individuals based on the CEU 1000 Genomes Phase 3 data 
using HapGen2. We selected a random sample of 2000 from this population and only retained 
the 334 variants with MAF > 0.005 in this sample. This genotype matrix is available in flashfm as the object X.

Measurements of two traits were generated using the mvrnorm function in R and genotype matrix X. In this example there are no missing
measurements, so N=2000 for both traits. 


```{r}
library(flashfm)
head(snpinfo)
covY
ybar
lapply(beta,head)
X[1:5,1:5]
N
head(raf)
```

### SNP correlation matrix and RAF vector provided - run single-trait and multi-trait fine-mapping 

This example takes seconds to run since the output is saved as an object, fm, in the flashfm package - details follow.

In this example, we make use of a SNP correlation matrix and reference allele frequency (RAF) vector as input to flashfm and use the internal 
flashfm function for single-trait fine-mappping, JAMexpandedCor.multi. This function is an expanded version of 
[JAM](https://onlinelibrary.wiley.com/doi/full/10.1002/gepi.21953). The function FLASHFMwithJAM provides a wrapper to run single-trait 
fine-mapping on each trait and then uses these results as input to flashfm. The SNP groups are also constructed within this wrapper function, 
as well as output from flashfm summarised at both the SNP-level and SNP group level.

We first need the correlation matrix, which we compute from our genotype matrix X, retaining only the SNPs that are in the raf vector; the raf 
vector was previously filtered at MAF 0.005.

```{r}
msnps <- names(raf) # filtered on maf 0.005
corX = cor(X[,msnps])
```

The command to run is as below, and for convenience the output from this function is saved in the object fm. 
In order to use JAM for single-trait fine-mapping, the R2BGLiMS library must be loaded, in addition to flashfm.

```{r,eval=FALSE} 
library(R2BGLiMS)
fm <- FLASHFMwithJAM(beta, corX, raf, ybar, N, r2 = 0.99, save.path="DIRtmp",TOdds=1,covY,cpp=0.99,NCORES=2) 
``` 

where arguments are described in detail here, though each object gives an example of format:

beta is a list of named vectors for each trait to be included in fine-mapping; see example format of object beta in this package;
create this object using b <- vector("list",2) and then assign beta vectors to each b[[i]], i=1,2.

corX is a SNP correlation matrix from an in-sample or reference panel; 
corX may have column and row names as in this example, though this is not necessary. Otherwise, column names are taken from the raf vector

raf is a reference allele frequency named vector with names given by snp IDs; snps must appear in same order as in corX; 
the reference allele must match the effect allele in the GWAS that generated the beta vectors

ybar is a vector of trait means; if traits transformed to normal(0,1), set ybar to a vector of 0.

N is a vector of sample sizes for traits, in same order as trait effect vectors given in beta. This can be based on data, if
unrelated samples, as here. Especially for samples with related individuals, we recommend finding the effective sample size for
each trait using the built-in function: 

Neff(raf, seB, Vy = 1), 

where raf is the reference allele frequency vector, seB is the vector of standard errors of the effect estimates in the same order 
of SNPs in raf, and Vy is the trait variance, which is 1 by default, assuming traits are transformed to normal(0,1). 

r2 is the level of tagging to use within JAM and r2=0.99 is recommended.

save.path lists a path to a directory to save the intermediate files generated by JAM. This directory can be created within R using the 
command dir.create(save.path).

TOdds=1 is the default for flashfm which gives a 50/50 odds that the traits share at least one causal variant

cpp=0.99 is the default for flashfm which prioritises single-trait fine-mapping models with cumulative posterior probability 0.99

NCORES is set to the number of traits to allow parallel processing in flashfm, though if on a Windows machine use NCORES=1

Our top group models are:

```{r}
fm$mpp.pp$PPg
```

Let’s check if our causal variants (in cvs vector) are included in the SNP groups from single and multi-trait fine-mapping and the sizes of 
the SNP groups

```{r}
flashfm:::groupIDs.fn(fm$snpGroups[[1]],cvs)
flashfm:::groupIDs.fn(fm$snpGroups[[2]],cvs)
fm$snpGroups$group.sizes
```

Notice that the groups for the causal variants that were not shared (cvs[c(2,3)]) are the same size for both methods and that for the shared 
variant (cvs[1]), flashfm constructs a smaller group than FINEMAP. This illustrates how sharing information between the traits can refine 
fine-mapping resolution.

### Reference genotype matrix provided

This example takes seconds to run.

In this example, we make use of a reference genotype matrix as input to flashfm and use the internal flashfm function for single-trait 
fine-mappping, JAMexpanded.multi. This function is an expanded version of 
[JAM](https://onlinelibrary.wiley.com/doi/full/10.1002/gepi.21953), and requires a pathway to the software 
[PLINK](http://zzz.bwh.harvard.edu/plink/).

For efficiency, we provide the output from JAMexpanded.multi in the object JAMmain.input. The command used to generate it is:

```{r,eval=FALSE}
JAMmain.input <- JAMexpanded.multi(beta,X,snpinfo,ybar,diag(covY),N,chr=10,fstub=fstub,mafthr=0.005,path2plink="/software/plink",r2=0.99,save.path=fstubJ,related=FALSE,y=NULL)
```

In the above command, ftsub and fstubJ are file prefixes, including file path, to save intermediate results. The related logical flag 
indicates whether or not there are related individuals in the sample. In the simulated data there is no relatedness, so relate=FALSE. If 
relatedness, then effective sample sizes need to be input for N; the built-in function Neff estimates effective sample size for each trait from 
the GWAS summary statistics.

Flashfm is then run with the following lines:

```{r}
ss.stats <- summaryStats(Xmat=TRUE,ybar.all=ybar,main.input=JAMmain.input)
fm.multi <- flashfm(JAMmain.input, TOdds=1,covY,ss.stats,cpp=0.99,maxmod=NULL,fastapprox=FALSE)
```

and SNP groups for both single-trait fine-mapping and flashfm are constructed by 

```{r} 
snpGroups <- makeSNPgroups2(JAMmain.input,fm.multi,is.snpmat=TRUE,r2.minmerge=0.7) 
``` 

where the is.snpmat flag=TRUE indicates that the reference genotype matrix was provided and 
r2.minmerge is the minimum LD between SNPs in different groups for them to merge; r2.minmerge=0.7 means 
that if a SNP in 1 group has r2>0.7 with a SNP in another group, these 2 groups are merged.
The first list of SNP groups were based on single-trait fine-mapping and the second list is based on flashfm.

The next function summarises the fine-mapping results using the SNP groups:

```{r}
mpp.pp <- PPsummarise(fm.multi, snpGroups, minPP=0.05)
```

This returns a list of 4 objects: MPP lists trait-specific PP of SNP inclusion in a model; MPPg lists trait-specific PP of SNP group 
inclusion in a model; PP lists trait-specific model PP; PPg lists trait-specific model PP in terms of  SNP group
Setting minPP=0.05 restricts output for group models to only those with PP>0.05 for at least one trait.

Our top group models are: 
```{r}
mpp.pp$PPg
```

Let's check if our causal variants (in cvs vector) are included in the SNP groups from single and multi-trait fine-mapping and the sizes 
of the SNP groups

```{r}
flashfm:::groupIDs.fn(snpGroups[[1]],cvs)
flashfm:::groupIDs.fn(snpGroups[[2]],cvs)
snpGroups$group.sizes
```
Notice that the groups for the causal variants that were not shared (cvs[c(2,3)]) are the same size for both methods and that
for the shared variant (cvs[1]), flashfm constructs a smaller group than JAMexpanded.multi. This illustrates how sharing information
between the traits can refine fine-mapping resolution.


### Reference genotype covariance matrix, RAF vector, and FINEMAP results for each trait provided 

This example takes around one minute to run.

Here, we use single-trait fine-mapping results from [FINEMAP](https://academic.oup.com/bioinformatics/article/32/10/1493/1743040) and 
show how to prepare it for use in flashfm.

The *.config files from FINEMAP are processed by the lines
```{r,eval=FALSE}
FMconfig <- vector("list",2)
for(i in 1:2) FMconfig[[i]] <- read.table(paste0("finemap",i,".config"), header = TRUE, as.is = TRUE, sep = " ")
```
and saved in the flashfm object FMconfig, which is prepared for flashfm by creating a list of the FINEMAP results for each trait:

```{r}
modPP.list <- vector("list",2)
for(i in 1:2) {modPP.list[[i]] <- data.frame(FMconfig[[i]][,c("config","prob")]); colnames(modPP.list[[i]]) <- c("snps","PP")}
names(modPP.list) <- names(beta)
lapply(modPP.list,head)
```

We provide the covariance matrix of the genotypes from X:
```{r}
msnps <- names(raf) # filtered on maf 0.005
covX = cov(X[,msnps])
```

However, if the SNP correlation matrix is available, the covariance matrix may be found using the correlation matrix and reference allele 
frequencies in the function cor2cov. This function is from the MBESS package and has been imported to flashfm; cor2cov is authored by Ken 
Kelley (University of Notre Dame; (\email{KKelley@@ND.Edu}) and Keke Lai (the \code{MBESS} package), with modifications by Dustin Fife 
\email{fife.dustin@@gmail.com}.

```{r,eval=FALSE}
covX <- cor2cov(corX,sd=sqrt(2*raf*(1-raf)))
```

N is a vector of sample sizes for traits, in same order as trait effect vectors given in beta. This can be based on data, if
unrelated samples, as here. Especially for samples with related individuals, we recommend finding the effective sample size for
each trait using the built-in function: 

Neff(raf, seB, Vy = 1), 

where raf is the reference allele frequency vector, seB is the vector of standard errors of the effect estimates in the same order 
of SNPs in raf, and Vy is the trait variance, which is 1 by default, assuming traits are transformed to normal(0,1). 


Similar lines to the previous example are run, but flashfm.input is needed to prepare input to flashfm, and its output has the
same form as JAMexpanded.multi.

```{r}
FMmain.input <- flashfm.input(modPP.list,beta1.list=beta,Gmat=covX,Nall=N,ybar.all=ybar,is.snpmat=FALSE,raf=raf)
ss.stats <- summaryStats(Xmat=FALSE,ybar.all=ybar,main.input=FMmain.input)
fm.multi <- flashfm(FMmain.input, TOdds=1,covY,ss.stats,cpp=0.99,maxmod=NULL,fastapprox=FALSE)
```

```{r}
snpGroups <- makeSNPgroups2(FMmain.input,fm.multi,is.snpmat=FALSE,r2.minmerge=0.7)
mpp.pp <- PPsummarise(fm.multi, snpGroups, minPP=0.05)
```

Our top group models are:
```{r}
mpp.pp$PPg
```

Let's check if our causal variants (in cvs vector) are included in the SNP groups from single and multi-trait fine-mapping and the sizes
of the SNP groups

```{r}
flashfm:::groupIDs.fn(snpGroups[[1]],cvs)
flashfm:::groupIDs.fn(snpGroups[[2]],cvs)
snpGroups$group.sizes
```

Notice that the groups for the causal variants that were not shared (cvs[c(2,3)]) are the same size for both methods and that
for the shared variant (cvs[1]), flashfm constructs a smaller group than FINEMAP.
This illustrates how sharing information between the traits can refine fine-mapping resolution.




